### 2019-07-10 2-Chapters/1-Chapter/chapter1.tex
Corrections fautes orthographes
Vérifications grâce à emccabedj.ucc.asn.au/checktex.html, avant intégration des commentaires d'Emilie sur ma conclusion

This third kind of wireless systems are characterized by their decentralized nature,
where communications are initialized and regulated by the devices, not by a centralized control systems.
% due to the really limited information that the base station can send to the devices (in downlink), it is no longer possible to optimize the efficiency of the network from the centralized point-of-view of the base station.
Indeed, a central control requires signaling packets that have been identified as too heavy for these kinds of systems.
In the present and future IoT networks, many devices of heterogeneous natures are using the same antenna for different applications.
A common problem is the strong constraint that such such IoT devices have on their power consumption, as most of them will be deployed without a direct power access and will run on a tiny battery, which lifespan should be maximized.
Most commercial IoT companies nowadays indeed sell a lifespan larger than $10$ years, like SIGFOX \cite{Centenaro16}.
Another common constraint for IoT devices is their low duty cycles, as most applications target a need for one or a few messages to send every day, in strike opposition to the high data-rate pursued for centralized systems (such as 4G/5G and WiFi).
%
Many different standards for IoT networks have been proposed in the recent years,
and they consist in a specification for both the \emph{PHY}sical
and the \emph{M}edium \emph{A}ccess \emph{C}ontrol (\emph{MAC}) layers.
To quote some examples of IoT standards, ZigBee, Z-Wave or Bluetooth are targeting short-range communications (up-to \SI{2}{\meter}), while LoRaWAN, SIGFOX, Ingenu or Weightless are designed for long-range communications (up-to \SI{50}{\kilo\meter}).
We refer to the survey \cite{Centenaro16} and references in our recent articles \cite{MoyBesson2019,MoyBesson2019Annales} for more details.

@@

\paragraph{From IT to Cognitive Radio.}
%
We can narrow down the first field of study of this thesis, steps by steps:
from Information Technologies (IT), to Telecommunications, to Wireless Communications, then to Cognitive Radio (CR),
and finally to Software Defined Radio (SDR).
%
The transition from the historical approach of hardware-based radio to SDR architectures is a gradual process, that started in the early $1990$s and has accelerated in the $2000$s.
% https://en.wikipedia.org/wiki/Software-defined_radio
A SDR is a radio communication system where components that have been traditionally implemented in hardware (\eg, mixers, filters, modulators/demodulators, detectors, etc) are more and more implemented by means of software on a processor.
%  with the exception of a unique pair of receiver/transmitter antennas.
Even though the SDR paradigm was initiated by the United States of North America defense research in the $1980$s, over the past $20$ years the civil industry has started to be interested by SDR, and CR has got significant attention from the academia as well.
% CR has been the first answer to solve spectrum scarcity issue which collected a huge attention in early 21st century.
%
As CR is not a standard technology, it does not have a single definition, so let us start by quoting the definition of two researchers whose works constitutes the roots of the development of CR, the first one having appeared just $20$ years ago.
% - technologies de l'information > télécommunications > communications sans fil > radio intelligente > DSA > OSA
%
% je peux faire comme Navik, donner des définitions de la CR, citant d'autres travaux
\begin{itemize}\tightlist
    \item
    Joseph Mittola in $1999$ proposed that
    \emph{``a really smart radio that would be self-, RF- and user-aware, and that would include software technology and machine learning capabilities along with a lot of high-fidelity knowledge of the radio environment''} \cite{Mitola99}.

    \item
    Then Simon Haykin in $2005$ also said that
    \emph{``a CR is an intelligent wireless communication system that is capable of being aware of its surroundings, learning, and adapting its operating parameters (\eg, transmit power and carrier frequency) on the fly with an objective of providing reliable anytime, anywhere, and spectrally efficient communication''} \cite{Haykin05}.
    % He was the first to propose to use CR for Dynamic Spectrum Access.

@@

% \paragraph{Chapter~\ref{chapter:25}.}
%
Because there are so many MAB algorithms available, we are also interested by the question of how an engineer can chose the algorithm to implement, in a given IoT object, in order to equip this object with the capacity to adapt robustly to any environment.
To answer this question, we present two approaches.
The first one is an empirical comparison of the most efficient and well known existing algorithms, in Section~\ref{sec:3:reviewSPAlgorithms} and \ref{sec:3:timeAndMemoryCosts}.
We confirm that widely used and not too sophisticated algorithms, such as \UCB{} \cite{Auer02}, Thompson sampling \cite{Thompson33} and \klUCB{} \cite{KLUCBJournal}, are the most efficient in terms of regret, and offer a good balance between regret and (time and memory) complexity.
The second approach is online algorithm selection, consisting in aggregating a (finite) set of algorithms and automatically discovering which one performs the most efficiently, against a given problem, with our contribution \Aggr{} that we detail in Chapter~\ref{chapter:25}.
This work on aggregating MAB algorithms was motivated for the OSA case of CR, where many previous research works only considered the $\UCB_1$ algorithm without really justifying their choice.
This contribution was presented at the IEEE WCNC conference in Barcelona, Spain, in April $2018$.

@@

% \paragraph{Chapter~\ref{chapter:4}.}
%
We study two models of IoT networks in Chapter~\ref{chapter:4}.
The first model considers IoT devices that simply have data to regularly send to a base station (at unpredictable or random times).
Such devices use acknowledgements as feedback, in order to optimize their up-link communications, by accessing more the best channels (\ie, the channel less occupied by the surrounding traffic).
The goal of this first model is to increase the Quality of Service (QoS) of such IoT network, by applying decentralized RL on the device side in order to reduce the rate of failed transmissions.
The base station will thus receive more up-link packets from the devices being served, if they can successfully learn by themselves an efficient spectrum access scheme.
It implies that the network can serve more end-devices while maintaining the same QoS.
%
Our second model then considers packet retransmissions, and while the two models are similar, the application of an efficient decentralized learning algorithm now results in both an increased battery life for each of the devices, as well as an increased QoS for the entire network, as less retransmissions reduce the local spectrum load.

@@

% We consider an ALOHA-based protocol, slotted in both time and frequency, by assuming a simple pre-agreed synchronization between all devices and the base station. This synchronization is the only aspect of the network that requires central control from the base station, and it only requires one downlink packet when a new device joins the network operated by this base station \cite{Centenaro16}.

@@

% The idea is that when a device sends an uplink message but does not receive its \Ack{} (before a certain delay), instead of going back to sleep mode in order to wait for its next activation (\eg, following the Bernoulli random activation pattern, or waiting for the application to need to send some data), the device will try again to send the same packet, a few times (\eg, up-to $m \leq 10$ times), after some small random waiting times.

@@

The second model with retransmissions concluded our collaboration with R. Bonnefoi, and resulted to a paper presented at the $1^{\text{st}}$ MoTION workshop, also during the IEEE WCNC $2019$ conference \cite{Bonnefoi2019WCNC}.

% \paragraph{Chapter~\ref{chapter:5}.}
%
In both models of IoT networks presented above, the core idea is to let every dynamic device run a learning algorithm in a fully decentralized way, in order to optimize the entire system.
It is important to note that each devices communicates, and thus learns, only at some rounds and not at all the time steps, by following its own random activation process (we restrict to a purely random Bernoulli activation process of probability $p$).
It means that each of them target its own local objective, which is to maximize its cumulated reward, in a selfish way and by being agnostic about the other devices following the same objective.
It is well known in game theory that playing selfishly can be disastrous for the centralized performance measure, one can think of popular ``dilemmas'', such as the \href{https://en.wikipedia.org/wiki/Prisoner%27s_dilemma}{prisoner dilemma}.
% , and for more examples we refer to the book \cite{CesaLugosi06}.
% https://www.ii.uni.wroc.pl/~lukstafi/pmwiki/uploads/AGT/Prediction_Learning_and_Games.pdf
So it was quite surprising that the numerical simulations, as well as the realistic PoC, showed that decentralized selfish MAB learning lead to efficient coordination between devices in all the considered scenarios, despite the fact that these IoT objects cannot communicate directly with each other, and receive only a collision feedback from the base station they are associated to.

@@

% Our work on multi-player bandit was presented at the ALT conference in Lanzarote, Spain, in April $2019$ \cite{Besson2018ALT},
% %
% and since then, some works have studied extensions of our framework or alternative algorithms, as a few independent teams started to work on different questions shortly after its publication.
% %
% Since last year, it was proven that the ``no sensing'' case (\ie, IoT) is essentially not harder that the ``sensing'' (\ie, OSA) case, from a theoretical point-of-view, by E. Boursier and V. Perchet \cite{BoursierPerchet18}.
% Even if their approach is costly empirically, the result is surprising and very interesting.
% They showed that even under the strong hypothesis of no direct communication between players and no central coordination (in the same model as the one of our paper), the players can actually use the collision feedback (\ie, receiving or not an \Ack) as an indirect way to communicate bits of information with each other.
% This idea requires that devices coordinate at least a little bit, by an initial pre-agreement (which is simply the fact that the IoT devices run the same algorithms).
% As they showed, the players can start by a ``musical chair'' phase, to reach an orthogonal allocation, and then they can alternate exploration, communication and exploitation phases until the end of the game at horizon $T$.
% Because they all follow the same algorithm, the devices are able to split the time steps between ``listening'' and ``talking'' phases, and the authors denote it the ``communication trick''.
% The communication trick is used by each player to send and receive observations about all arms from the other players, using one forced collision to send one bit at a time.
% They use about $T-\bigO{\log(T)}$ rounds in exploitation of the best arms, and only $\bigO{\log(T)}$ rounds in exploration and communication, yielding a logarithmic regret.
% % As they showed, after a long enough initialization period, the $M$ players have elected a ``master'' player and $M-1$ ``followers'' players (with high probability),
% Another direction, as followed by \cite{Bistritz18}, is to use this communication trick and reduce the problem to a centralized multiple-play (MP) MAB, which can be solved efficiently with the MP generalization of \klUCB{} \cite{Luedtke16} or Thompson sampling \cite{Komiyama15}.

@@

\item
    \emph{MALIN: ``Multi-Arm bandit Learning for Iot Networks'' with GRC: A TestBed Implementation and Demonstration that Learning Helps},
    by \textbf{Lilian Besson}, Rémi Bonnefoi, Christophe Moy.
    Demonstration presented at \emph{International Conference on Communication},
    % \footnote{~ICT 2018, see \href{http://ict-2018.org/demos}{\texttt{ict-2018.org/demos}}},
    Saint-Malo, France in June $2018$.
    See \href{https://YouTu.be/HospLNQhcMk}{\texttt{YouTu.be/HospLNQhcMk}} for a $6$-minutes presentation video.
    See Section~\ref{sec:4:gnuradio}.
    \cite{Besson2018ICT}

>>>>

This third kind of wireless systems are characterized by their decentralized nature,
where communications are initialized and regulated by the devices, not by a centralized control systems.
% due to the really limited information that the base station can send to the devices (in down-link), it is no longer possible to optimize the efficiency of the network from the centralized point-of-view of the base station.
Indeed, a central control requires signaling packets that have been identified as too heavy for these kinds of systems.
In the present and future IoT networks, many devices of heterogeneous natures are using the same antenna for different applications.
A common problem is the strong constraint that such IoT devices have on their power consumption, as most of them will be deployed without a direct power access and will run on a tiny battery, which lifespan should be maximized.
Most commercial IoT companies nowadays indeed sell a lifespan larger than $10$ years, like SIGFOX \cite{Centenaro16}.
Another common constraint for IoT devices is their low duty cycles, as most applications target a need for one or a few messages to send every day, in strike opposition to the high data-rate pursued for centralized systems (such as 4G/5G and WiFi).
%
Many different standards for IoT networks have been proposed in the recent years,
and they consist in a specification for both the \emph{PHY}sical
and the \emph{M}edium \emph{A}ccess \emph{C}ontrol (\emph{MAC}) layers.
To quote some examples of IoT standards, ZigBee, Z-Wave or Bluetooth are targeting short-range communications (up-to \SI{2}{\meter}), while LoRaWAN, SIGFOX, Ingenu or Weightless are designed for long-range communications (up-to \SI{50}{\kilo\meter}).
We refer to the survey \cite{Centenaro16} and references in our recent articles \cite{MoyBesson2019,MoyBesson2019Annales} for more details.

@@

\paragraph{From IT to Cognitive Radio.}
%
We can narrow down the first field of study of this thesis, steps by steps:
from Information Technologies (IT), to Telecommunications, to Wireless Communications, then to Cognitive Radio (CR),
and finally to Software Defined Radio (SDR).
%
The transition from the historical approach of hardware-based radio to SDR architectures is a gradual process, that started in the early $1990$s and has accelerated in the $2000$s.
% https://en.wikipedia.org/wiki/Software-defined_radio
A SDR is a radio communication system where components that have been traditionally implemented in hardware (\eg, mixers, filters, modulators/demodulators, detectors, etc) are more and more implemented by means of software on a processor.
%  with the exception of a unique pair of receiver/transmitter antennas.
Even though the SDR paradigm was initiated by the United States of North America defense research in the $1980$s, over the past $20$ years the civil industry has started to be interested by SDR, and CR has got significant attention from the academia as well.
% CR has been the first answer to solve spectrum scarcity issue which collected a huge attention in early 21st century.
%
As CR is not a standard technology, it does not have a single definition, so let us start by quoting the definition of two researchers whose works constitutes the roots of the development of CR, the first one having appeared just $20$ years ago.
% - technologies de l'information > télécommunications > communications sans fil > radio intelligente > DSA > OSA
%
% je peux faire comme Navik, donner des définitions de la CR, citant d'autres travaux
\begin{itemize}\tightlist
    \item
    J. Mittola in $1999$ proposed that
    \emph{``a really smart radio that would be self-, RF- and user-aware, and that would include software technology and machine learning capabilities along with a lot of high-fidelity knowledge of the radio environment''} \cite{Mitola99}.

    \item
    Then S. Haykin in $2005$ also said that
    \emph{``a CR is an intelligent wireless communication system that is capable of being aware of its surroundings, learning, and adapting its operating parameters (\eg, transmit power and carrier frequency) on the fly with an objective of providing reliable anytime, anywhere, and spectrally efficient communication''} \cite{Haykin05}.
    % He was the first to propose to use CR for Dynamic Spectrum Access.

@@

% \paragraph{Chapter~\ref{chapter:25}.}
%
Because there are so many MAB algorithms available, we are also interested by the question of how an engineer can choose the algorithm to implement, in a given IoT object, in order to equip this object with the capacity to adapt robustly to any environment.
To answer this question, we present two approaches.
The first one is an empirical comparison of the most efficient and well known existing algorithms, in Section~\ref{sec:3:reviewSPAlgorithms} and \ref{sec:3:timeAndMemoryCosts}.
We confirm that widely used and not too sophisticated algorithms, such as \UCB{} \cite{Auer02}, Thompson sampling \cite{Thompson33} and \klUCB{} \cite{KLUCBJournal}, are the most efficient in terms of regret, and offer a good balance between regret and (time and memory) complexity.
The second approach is online algorithm selection, consisting in aggregating a (finite) set of algorithms and automatically discovering which one performs the most efficiently, against a given problem, with our contribution \Aggr{} that we detail in Chapter~\ref{chapter:25}.
This work on aggregating MAB algorithms was motivated for the OSA case of CR, where many previous research works only considered the $\UCB_1$ algorithm without really justifying their choice.
This contribution was presented in the IEEE WCNC conference in Barcelona, Spain, in April $2018$.

@@

% \paragraph{Chapter~\ref{chapter:4}.}
%
We study two models of IoT networks in Chapter~\ref{chapter:4}.
The first model considers IoT devices that simply have data to regularly send to a base station (at unpredictable or random times).
Such devices use acknowledgements as feedback, in order to optimize their up-link communications, by accessing more the best channels (\ie, the channel less occupied by the surrounding traffic).
The goal of this first model is to increase the Quality of Service (QoS) of such IoT network, by applying decentralized RL on the device side in order to reduce the rate of failed transmissions.
The base station will thus receive more up-link packets from the devices being served, if they can successfully learn by themselves an efficient spectrum access scheme.
It implies that the network can serve more end-devices while maintaining the same QoS.
%
Our second model then considers packet retransmissions, and while the two models are similar, the application of an efficient decentralized learning algorithm now results in both an increased battery life for each of the devices, and an increased QoS for the entire network, as less retransmissions reduce the local spectrum load.

@@

% We consider an ALOHA-based protocol, slotted in both time and frequency, by assuming a simple pre-agreed synchronization between all devices and the base station. This synchronization is the only aspect of the network that requires central control from the base station, and it only requires one down-link packet when a new device joins the network operated by this base station \cite{Centenaro16}.

@@

% The idea is that when a device sends an up-link message but does not receive its \Ack{} (before a certain delay), instead of going back to sleep mode in order to wait for its next activation (\eg, following the Bernoulli random activation pattern, or waiting for the application to need to send some data), the device will try again to send the same packet, a few times (\eg, up-to $m \leq 10$ times), after some small random waiting times.

@@

The second model with retransmissions concluded our collaboration with R. Bonnefoi, and resulted to a paper presented in the $1^{\text{st}}$ MoTION workshop, also during the IEEE WCNC $2019$ conference \cite{Bonnefoi2019WCNC}.

% \paragraph{Chapter~\ref{chapter:5}.}
%
In both models of IoT networks presented above, the core idea is to let every dynamic device run a learning algorithm in a fully decentralized way, in order to optimize the entire system.
We highlight that each devices communicates, and thus learns, only at some rounds and not at all the time steps, by following its own random activation process (we restrict to a purely random Bernoulli activation process of probability $p$).
This means that each of them target its own local objective, which is to maximize its cumulated reward, in a selfish way and by being agnostic about the other devices following the same objective.
It is well known in game theory that playing selfishly can be disastrous for the centralized performance measure, one can think of popular ``dilemmas'', such as the \href{https://en.wikipedia.org/wiki/Prisoner%27s_dilemma}{prisoner dilemma}.
% , and for more examples we refer to the book \cite{CesaLugosi06}.
% https://www.ii.uni.wroc.pl/~lukstafi/pmwiki/uploads/AGT/Prediction_Learning_and_Games.pdf
So it was quite surprising that the numerical simulations, as well as the realistic PoC, showed that decentralized selfish MAB learning lead to efficient coordination between devices in all the considered scenarios, despite the fact that these IoT objects cannot communicate directly with each other, and receive only a collision feedback from the base station they are associated to.

@@

% Our work on multi-player bandit was presented in the ALT conference in Lanzarote, Spain, in April $2019$ \cite{Besson2018ALT},
% %
% and since then, some works have studied extensions of our framework or alternative algorithms, as a few independent teams started to work on different questions shortly after its publication.
% %
% Since last year, it was proven that the ``no sensing'' case (\ie, IoT) is essentially not harder that the ``sensing'' (\ie, OSA) case, from a theoretical point-of-view, by E. Boursier and V. Perchet \cite{BoursierPerchet18}.
% Even if their approach is costly empirically, the result is surprising and very interesting.
% They showed that even under the strong hypothesis of no direct communication between players and no central coordination (in the same model as the one of our paper), the players can actually use the collision feedback (\ie, receiving or not an \Ack) as an indirect way to communicate bits of information with each other.
% This idea requires that devices coordinate at least a little bit, by an initial pre-agreement (which is simply the fact that the IoT devices run the same algorithms).
% As they showed, the players can start by a ``musical chair'' phase, to reach an orthogonal allocation, and then they can alternate exploration, communication and exploitation phases until the end of the game at horizon $T$.
% Because they all follow the same algorithm, the devices are able to split the time steps between ``listening'' and ``talking'' phases, and the authors denote it the ``communication trick''.
% The communication trick is used by each player to send and receive observations about all arms from the other players, using one forced collision to send one bit at a time.
% They use about $T-\bigO{\log(T)}$ rounds in exploitation of the best arms, and only $\bigO{\log(T)}$ rounds in exploration and communication, yielding a logarithmic regret.
% % As they showed, after a long enough initialization period, the $M$ players have elected a ``master'' player and $M-1$ ``followers'' players (with high probability),
% Another direction, as followed by \cite{Bistritz18}, is to use this communication trick and reduce the problem to a centralized multiple-play (MP) MAB, which can be solved efficiently with the MP generalization of \klUCB{} \cite{Luedtke16} or Thompson sampling \cite{Komiyama15}.

@@

\item
    \emph{MALIN: ``Multi-Arm bandit Learning for Iot Networks'' with GRC: A TestBed Implementation and Demonstration that Learning Helps},
    by \textbf{Lilian Besson}, Rémi Bonnefoi, Christophe Moy.
    Demonstration presented in \emph{International Conference on Communication},
    % \footnote{~ICT 2018, see \href{http://ict-2018.org/demos}{\texttt{ict-2018.org/demos}}},
    Saint-Malo, France in June $2018$.
    See \href{https://YouTu.be/HospLNQhcMk}{\texttt{YouTu.be/HospLNQhcMk}} for a $6$-minutes presentation video.
    See Section~\ref{sec:4:gnuradio}.
    \cite{Besson2018ICT}
